Future Evolution of CSMA Protocols for the IEEE 802.11 Standard

### 2013-03-06 paper.tex
make differences
Azadeh s changes

\newcommand{\Az}[1]{{\bf{#1}}}

@@

A suitable candidate for the replacement of CSMA/CA, should
\begin{itemize}
  \item provide performance advantages in terms of throughput and/or short-term fairness;
  \item be backward compatible with the current implementation;
  \item support a large number of simultaneous contenders; and 
  \item be a simple evolution, in terms of implementation, to ease the transition and reduce the time to market (optional but desirable).
\end{itemize}

@@

When it comes to backward compatibility, neither of the aforementioned solutions are able to fairly share the medium with legacy devices.
In fact, these proposals are, generally speaking, less aggressive than the currently implemented CSMA/CA protocol. Consequently, in a hypothetical mixed network in which the new and old protocols coexisted, the new stations would receive a smaller share of the available bandwidth.

@@

The third group of solutions, which is the focus of the current paper, can deliver a throughput above the maximum attainable by CSMA/CA. This performance boost is achieved mainly by the use of deterministic backoff after successful transmissions, which reduces the chances of collisions for nodes that were successful in their previous transmission. Furthermore, under certain conditions, a collision-free operation can be reached. {We will refer hereafter to the class of algorithms that use deterministic backoffs after successful transmissions as CSMA with Enhanced Collision Avoidance (CSMA/ECA).} This approach was first introduced in \cite{barcelo2008lba}, and later a more detailed analysis for both saturated and non-saturated conditions was presented in \cite{bellalta2009vtc}.
A more in-depth study is carried out in \cite{he2009srb}, including realistic elements such as the possibility of Clear Channel Assessment errors.
Different aspects of fairness are addressed in \cite{he2009srb,barcelo2010fcc,fang2011dlm}.
The performance in realistic channels, taking into account the Auto Rate Fallback mechanism, is evaluated in \cite{martorell2012pec,martorell2012tfl}.

Quality-of-service and traffic differentiation in CSMA/ECA has is discussed in \cite{barcelo2009tpc, he2009srb}.
The performance of CSMA/ECA in the presence of channel errors is considered in \cite{he2009srb, barcelo2010fcc, fang2011dlm, barcelo2012mdc}. 
Other aspects such as coexistence with legacy stations and performance in the presence of new entrants have also been covered in \cite{he2009srb, barcelo2010fcc, fang2011dlm}.
Results for the impact of slot drift (non-ideal clocks) in protocols similar to CSMA/ECA can be found in \cite{gong2012asd}.
In the present workshop paper, we focus on a simple greenfield scenario with a single traffic class, ideal channel, ideal clocks, and saturation conditions.
We are currently evaluating more complex scenarios that we outline in Section \ref{sec:future_work}

@@

The same principles that we exploit to prevent collisions in WLANs can be used in other areas of radio resource management in wireless area networks \cite{duffy2011dcs,checco2012scs,checco2012lbc,khan2013aso}.

@@

Note that the coexistence and fairness of basic CSMA/ECA and CSMA/CA has been studied in \cite{barcelo2010fcc}. 
The addition of hysteresis and fair-share to the basic CSMA/ECA may represent less frequent transmissions and more packets per transmission
These two modifications do not interfere in any way with the execution of the legacy protocol and, therefore, CSMA/CA with hysteresis and fair-share can also coexist with the currently deployed protocol.

@@

Algorithm \ref{alg:CSMA_CA} describes the CSMA/CA algorithm that is used in current networks.
When a station joins the contention, it initializes the retry attempt counter $r$ and the backoff stage $s$ to zero. The backoff counter $b$ is initialized using a uniform random distribution and the minimum contention window $\text{CW}_{\min}$. After each collision, the retry attempt counter and the backoff stage counter are incremented. As a consequence of the incremented backoff stage, a larger contention window is used. Note that there is a maximum backoff stage $S$ and a maximum retry limit $R$ specified by the protocol. When the number of transmission attempts on a packet reaches the maximum retry limit, the packet is discarded. Furthermore, $r$ and $s$ are reset to zero, and a new value for $b$ is computed both when the retry limit is reached and when the transmission is successful. Note that the pseudo-code performs the same action in lines $18$ and $21$. Therefore, the algorithm could have been further simplified if instead this action were moved to immediately after line $16$. However, the current presentation of the algorithm eases the comparison between the CSMA/CA and the enhancements that follow.

@@

There are three changes in the CSMA/ECA with hysteresis and fair-share compared to the legacy CSMA/CA protocol. Firstly, a deterministic backoff is used after successful transmissions. This simple modification converts CSMA/CA into CSMA/ECA.
Secondly, the backoff stage is not reset after a packet is serviced.
The backoff stage is reset only when the station leaves the contention because it has no packet to be transmitted. This modification is what we have been referring to as hysteresis.
And thirdly, the number of packets transmitted in every transmission attempt is chosen as a function of the backoff stage, which is what we call fair-share.
Note that current standards already support the transmission of multiple packets in a single slot.

Algorithm \ref{alg:CSMA_ECA} describes the basic CSMA/ECA in which a deterministic backoff is used after successful transmissions. The only change with respect to CSMA/CA is in fact in line $18$, where the random assignment of $b$ in CSMA/CA is replaced by a deterministic assignment in CSMA/ECA. Note that at this point the value of $s$ is zero, and therefore the assigned deterministic value is in fact $b = \text{CW}_{\min}/2$.
This value is roughly equal to the expectation of the backoff chosen in line $18$ in Algorithm \ref{alg:CSMA_CA}.
This particular choice guarantees fairness between new stations and legacy stations.

@@

\begin{algorithm}[ht!!!]
\While{the device is on}
{
  %\tcc{Initialize retransmission attempts.}
  $r \leftarrow 0$ ; $s \leftarrow 0$ \;
  %\tcc{Initialize backoff counter.}
  $b \leftarrow \mathcal{U}[0,2^s\rm{CW}_{min}-1]$\;
  \While{there is a packet to transmit}{
    %\tcc{Initialize $a$.}
    \Repeat{($r = R$) or (success)}{
      %\tcc{First, backoff.}
      \While{$b>0$}{
        wait 1 slot \;
        $b \leftarrow b-1$ \;
      }
      \colorbox{yellow}{Attempt transmission of 1 packet } \;
      \If{collision}{
        %\tcc{Random backoff.}
        $r \leftarrow r+1$ \;
        $s \leftarrow \min (s+1,S)$ \;
        $b \leftarrow \mathcal{U}[0, 2^s \rm{CW}_{min} -1]$\;
      }
    }
    $r \leftarrow 0$ \;
    \colorbox{yellow}{$s \leftarrow 0$ } \;
	 \eIf{success}{
      %\tcc{Random backoff.}
      \colorbox{yellow}{$b \leftarrow \mathcal{U}[0,2^{s}\rm{CW}_{min}-1]$} \;
    }
    {
      Discard packet\;
      $b \leftarrow \mathcal{U}[0,2^s \rm{CW}_{min}-1]$\;
    }
  }
  Wait until there is a packet to transmit \;
}
\caption{CSMA/CA}
\label{alg:CSMA_CA}
\end{algorithm}

\begin{algorithm}[ht!!!]
\While{the device is on}
{
  $r \leftarrow 0$ ; $s \leftarrow 0$ \;
  $b \leftarrow \mathcal{U}[0,2^s\rm{CW}_{min}-1]$\;
  \While{there is a packet to transmit}{
    %\tcc{Initialize $a$.}
    \Repeat{($r = R$) or (success)}{
      %\tcc{First, backoff.}
      \While{$b>0$}{
        wait 1 slot \;
        $b \leftarrow b-1$ \;
      }
      \colorbox{yellow}{Attempt transmission of 1 packet } \;
      \If{collision}
      {
        %\tcc{Random backoff.}
        $r \leftarrow r+1$ \;
        $s \leftarrow \min (s+1,S)$ \;
        $b \leftarrow \mathcal{U}[0, 2^s  \rm{CW}_{min} -1]$\;
      }
    }
    $r \leftarrow 0$ \;
    \colorbox{yellow}{$s \leftarrow 0$} \;    
    \eIf{success}{
      %\tcc{Random backoff.}
      \colorbox{yellow}{$b \leftarrow (2^{s}\rm{CW}_{min})/2$} \;
    }
    {
      Discard packet\;
      $b \leftarrow \mathcal{U}[0, 2^s \rm{CW}_{min}-1]$\;
    }
  }
  Wait until there is a packet to transmit \;
}
\caption{CSMA/ECA}
\label{alg:CSMA_ECA}
\end{algorithm}

\begin{algorithm}[ht!!!]
\While{the device is on}
{
  $r \leftarrow 0$ ; $s \leftarrow 0$ \;
  $b \leftarrow \mathcal{U}[0,2^s\rm{CW}_{min}-1]$\;
  \While{there is a packet to transmit}{
    %\tcc{Initialize $a$.}
    \Repeat{($r = R$) or (success)}{
      %\tcc{First, backoff.}
      \While{$b>0$}{
        wait 1 slot \;
        $b \leftarrow b-1$ \;
      }
      \colorbox{yellow}{Attempt transmission of 1 packet } \;
      \If{collision}{
        %\tcc{Random backoff.}
        $r \leftarrow r+1$ \;
        $s \leftarrow \min (s+1,S)$ \;
        $b \leftarrow \mathcal{U}[0, 2^s  \rm{CW}_{min} -1]$\;
      }
    }
    $r \leftarrow 0$ \;
    %$s \leftarrow 0$ \;
    \eIf{success}{
      %\tcc{Random backoff.}
      \colorbox{yellow}{$b \leftarrow (2^{s}\rm{CW}_{min})/2$} \;
    }
    {
      Discard packet\;
      $b \leftarrow \mathcal{U}[0, 2^s \rm{CW}_{min}-1]$\;
    }
  }
  Wait until there is a packet to transmit \;
}
\vspace{0.2cm}
\caption{CSMA/ECA with hysteresis}
\label{alg:CSMA_ECA_hist}
\end{algorithm}

\begin{algorithm}[ht!!!]
\While{the device is on}
{
  $r \leftarrow 0$ ; $s \leftarrow 0$ \;
  $b \leftarrow \mathcal{U}[0,2^s\rm{CW}_{min}-1]$\;
  \While{there is a packet to transmit}{
    %\tcc{Initialize $a$.}
    \Repeat{($r = R$) or (success)}{
      %\tcc{First, backoff.}
      \While{$b>0$}{
        wait 1 slot \;
        $b \leftarrow b-1$ \;
      }
      \colorbox{yellow}{Attempt transmission of $2^s$ packets } \;
      \If{collision}{
        %\tcc{Random backoff.}
        $r \leftarrow r+1$ \;
        $s \leftarrow \min (s+1,S)$ \;
        $b \leftarrow \mathcal{U}[0, 2^s  \rm{CW}_{min} -1]$\;
      }
    }
    $r \leftarrow 0$ \;
    %$s \leftarrow 0$ \;
    \eIf{success}{
      %\tcc{Random backoff.}
      \colorbox{yellow}{$b \leftarrow (2^{s}\rm{CW}_{min})/2$} \;
    }
    {
      Discard packet\;
      $b \leftarrow \mathcal{U}[0, 2^s \rm{CW}_{min}-1]$\;
    }
  }
  Wait until there is a packet to transmit \;
}
\vspace{0.2cm}
\caption{CSMA/ECA with hysteresis and fair-share}
\label{alg:CSMA_ECA_hist_fair}
\end{algorithm}

An example of the operation of CSMA/ECA with hysteresis is presented in Fig. \ref{fig:csma_eca_different_backoff}.

\begin{figure*}[!t]
\centering
\includegraphics[width=7.0in]{figures/csma_eca_different_backoff}
\caption{Schedule length distributed adaption example. All the stations transmit the same number of packets in each cycle, despite using different schedule lengths.}
\label{fig:csma_eca_different_backoff}
\end{figure*}

@@

In this section, the performance of the different CSMA protocols is evaluated through simulations. The presented results show the saturation throughput and fairness (Jain's Fairness Index (JFI) \cite{jain1984quantitative}) when the number of contenders increases. Additionally, the evolution of the probabilities that one slot collision is plotted to provide further insights on the operation of each protocol, and specially whether the collision-free operation is reached.

@@

To evaluate the four protocols, a network of $N$ nodes is considered, where each node is within the coverage area of the others. The channel does not introduce errors and the nodes are set to be under saturation (always have packets to transmit). The number of contenders, $N$, ranges from $2$ to $50$, and one thousand instances of the simulation are performed for each $N$. All plotted results show the 95\% confidence intervals.

@@

In Figure \ref{Fig:throughput}, the throughput of all four schemes is plotted. As can be observed, the throughput for CSMA/CA decreases with the number of nodes, as the binary exponential backoff reduces the number of transmission attempts to keep the collisions low. The basic CSMA/ECA is able to achieve the collision-free operation if the number of nodes is lower than the deterministic backoff value, i.e., $N\leq \text{CW}_{\min}/2$. This is why in this figure, a phase transition is observed at $N=8$. When $N$ is larger than $\text{CW}_{\min}/2=8$, although the collision-free operation is not possible, the throughput obtained remains higher than in CSMA/CA. This is because, regardless of whether or not the collision-free operation is reached, in CSMA/ECA, a node always has a lower collision probability immediately after a successful transmission, without affecting the collision probability of other nodes.

@@

Clearly, reaching a collision-free operation is highly desirable, but not always possible in the basic CSMA/ECA. CSMA/ECA with Hysteresis exactly addresses this issue by simply not resetting the backoff stage after a successful transmission. In other words, the deterministic backoff value ($2^s \text{CW}_{\min}/2$) is chosen based on the backoff stage, $s$, in which the transmission was successful. This way, a larger number of contenders can reach a collision-free operation (Figure \ref{Fig:throughput}), but at the expense of a lower throughput for $N\leq 8$ compared to the basic CSMA/ECA. This is because if collision-free operation can be reached with a lower deterministic backoff value, there will be fewer empty slots, hence higher channel access efficiency. Furthermore, CSMA/ECA with hysteresis has a lower long-term fairness (Figure \ref{Fig:fairness}) compared to the basic CSMA/ECA or even the legacy CSMA/CA protocol, as once the network has reached the collision-free state, nodes that use a large backoff stage will have fewer transmission opportunities than others.

To address the aforementioned fairness issue when hysteresis is used, CSMA/ECA with hysteresis and fair-share was introduced. It achieves greater throughput than CSMA/ECA with hysteresis only, while being fair for any number of contenders. This is achieved by using packet aggregation, transmitting $2^s$ packets per attempt. As observed in Figure \ref{Fig:throughput}, using aggregation we are not only able to provide a fair access, but also we are able to significantly increase the throughput. {Note that the throughput increases with the number of nodes because for larger $N$, the collision-free operation is reached at higher backoff stages, which means more packets will be aggregated and transmitted at every attempt, thus improving the efficiency of the network.}

@@

\section{Future Work}
\label{sec:future_work}

In this workshop paper, we have evaluated the protocols under discussion in a very simple scenario using ideal channel, ideal clocks and saturation conditions.
Furthermore, all the participating stations use the same protocol and have a single traffic  class.
The presented metrics include only throughput and fairness, whereas delay, jitter and packet have been omitted.
Another scenario of interest that we have not included is the situation in which the stations join the contention to transmit a file and then they leave.

Our proposed protocol is based on CSMA/ECA, which works well in non-ideal conditions, and the results we are currently obtaining confirm that CSMA/ECA with hysteresis and fair-share inherits the flexibility and robustness of its predecessors.

Regarding the implementation in real prototypes, it has been proven difficult until now, as the IEEE 802.11 protocols are embedded in the network interfaces and it is not possible to reprogram them.
There are current advances in that direction \cite{tinnirello2012wmp} but it is still difficult to set the backoff to a deterministic value after successful transmissions in available hardware.

Our intention is to try our protocols in a different technology, namely radio-frequency identification (RFID), where there is programmable hardware that can be used to implement arbitrary protocols.
RFID uses also a contention protocol which is prone to collisions, and we believe it can be a first area in which we will be available to develop working prototypes to validate some of the ideas proposed in the present paper.

\section{Conclusion}

In this paper we have described a suitable replacement of the prevalent CSMA/CA protocol used in the IEEE 802.11 standard. Compared to the CSMA/CA protocol, the proposed protocol, called CSMA/ECA with Hysteresis and Fair-Share, offers performance boost in terms of throughput while preserving long-term fairness. In fact, under certain conditions, collision-free operation can be reached in a distributed and adaptive manner, even as the number of nodes in the network increases. Furthermore, this protocol is backward compatible, and can operate with nodes using the legacy protocol without degrading the performance of the legacy nodes. All these properties make the proposed protocol a good candidate to replace CSMA/CA in the upcoming revisions of the standard.

>>>>

%\newcommand{\Az}[1]{{\bf{#1}}}
%\newcommand{\Az}[1]{{\color{blue}{#1}}}
\newcommand{\Az}[1]{{\color{black}{#1}}}
%\newcommand{\AzCom}[1]{{\it \color{magenta} [#1]}}
\newcommand{\AzCom}[1]{}
%\newcommand{\AzDel}[1]{{\color{red}{#1}}}
\newcommand{\AzDel}[1]{}

@@

A suitable candidate to replace CSMA/CA, should
\begin{itemize}
  \item provide performance advantages in terms of throughput and/or short-term fairness;
  \item be backward compatible with the current implementation;
  \item support a large number of simultaneous contenders; and
  \item be a simple evolution, in terms of implementation, to ease the transition and reduce the time to market (optional but desirable).
\end{itemize}

@@

When it comes to backward compatibility, neither of the aforementioned solutions are able to fairly share the medium with legacy devices.
In fact, these proposals are, generally speaking, less aggressive than the currently implemented CSMA/CA protocol. Consequently, in a hypothetical mixed network in which the new and old protocols coexist, the new stations receive a smaller share of the available bandwidth.

@@

The third group of solutions, which is the focus of the current paper, can deliver a throughput above the maximum attainable by CSMA/CA. This performance boost is achieved mainly by the use of deterministic backoff after successful transmissions, which reduces the chances of collisions for nodes that were successful in their previous transmission. Furthermore, under certain conditions, a collision-free operation can be reached. {We will refer hereafter to the class of algorithms that use deterministic backoffs after successful transmissions as CSMA with Enhanced Collision Avoidance (CSMA/ECA).} This approach was first introduced in \cite{barcelo2008lba}, and later a more detailed analysis for both saturated and unsaturated conditions was presented in \cite{bellalta2009vtc}.
A more in-depth study is carried out in \cite{he2009srb}, including realistic elements such as the possibility of Clear Channel Assessment errors.
Different aspects of fairness are addressed in \cite{he2009srb,barcelo2010fcc,fang2011dlm}.
The performance in realistic channels, taking into account the Auto Rate Fallback mechanism, is evaluated in \cite{martorell2012pec,martorell2012tfl}.

Quality-of-service and traffic differentiation in CSMA/ECA is discussed in \cite{barcelo2009tpc, he2009srb}.
The performance of CSMA/ECA in the presence of channel errors is considered in \cite{he2009srb, barcelo2010fcc, fang2011dlm, barcelo2012mdc}.
Other aspects such as coexistence with legacy stations and the performance when nodes can arbitrarily enter and leave the network have been covered in \cite{he2009srb, barcelo2010fcc, fang2011dlm}.
Results for the impact of slot drift (non-ideal clocks) in protocols similar to CSMA/ECA can be found in \cite{gong2012asd}.

@@

\AzCom{Greenfield paragraph moved to the Results section} The same principles that we exploit to prevent collisions in WLANs can be used in other areas of radio resource management in wireless area networks \cite{duffy2011dcs,checco2012scs,checco2012lbc,khan2013aso}.

@@

Note that the coexistence and fairness of basic CSMA/ECA and CSMA/CA has been studied in \cite{barcelo2010fcc}.
The addition of hysteresis and fair-share to the basic CSMA/ECA may represent less frequent transmissions and more packets per transmission
These two modifications do not interfere in any way with the execution of the legacy protocol; therefore, CSMA/CA with hysteresis and fair-share can also coexist with the currently deployed protocol.

@@

Algorithm \ref{alg:CSMA_CA} describes the CSMA/CA protocol that is used in current networks.
When a station joins the contention, it initializes the retry attempt counter $r$ and the backoff stage $s$ to zero. The backoff counter $b$ is initialized using a uniform random distribution and the minimum contention window $\text{CW}_{\min}$. After each collision, the retry attempt counter and the backoff stage counter are incremented. As a consequence of the incremented backoff stage, a larger contention window is used. Note that there is a maximum backoff stage $S$ and a maximum retry limit $R$ specified by the protocol. When the number of transmission attempts on a packet reaches the maximum retry limit, the packet is discarded. Furthermore, $r$ and $s$ are reset to zero, and a new value for $b$ is computed both when the retry limit is reached and when the transmission is successful. Note that the pseudo-code performs the same action in lines $18$ and $21$. Therefore, the algorithm could be further simplified if instead this action were moved to immediately after line $16$. However, the current presentation of the algorithm eases the comparison between the CSMA/CA and the enhancements that follow.

@@

There are three changes in the CSMA/ECA with hysteresis and fair-share compared to the legacy CSMA/CA protocol. Firstly, a deterministic backoff is used after successful transmissions. This simple modification converts CSMA/CA into basic CSMA/ECA.
Secondly, the backoff stage is not reset after a packet is serviced.
The backoff stage is reset only when the station leaves the contention because it has no packet to be transmitted. This modification is what we have been referring to as hysteresis.
And thirdly, the number of packets transmitted in every transmission attempt is chosen as a function of the backoff stage, which is what we call fair-share.
Note that current standards already support the transmission of multiple packets in a single slot.

Algorithm \ref{alg:CSMA_ECA} describes the basic CSMA/ECA in which a deterministic backoff is used after successful transmissions. The only change with respect to CSMA/CA is in fact in line $18$, where the random assignment of $b$ in CSMA/CA is replaced by a deterministic assignment in CSMA/ECA. Note that at this point the value of $s$ is zero, and therefore the assigned deterministic value is in fact $b = \text{CW}_{\min}/2-1$. This value is roughly equal to the expectation of the backoff chosen in line $18$ in Algorithm \ref{alg:CSMA_CA}. This particular choice improves fairness between new stations and legacy stations. \AzCom{Does it? Even if they are not using aggregation?}

\begin{algorithm}[ht!!!]
\While{the device is on}
{
  %\tcc{Initialize retransmission attempts.}
  $r \leftarrow 0$; $s \leftarrow 0$\;
  %\tcc{Initialize backoff counter.}
  $b \leftarrow \mathcal{U}[0,2^s{\rm{CW}_{min}}-1]$\;
  \While{there is a packet to transmit}{
    %\tcc{Initialize $a$.}
    \Repeat{($r = R$) or (success)}{
      %\tcc{First, backoff.}
      \While{$b>0$}{
        wait 1 slot\;
        $b \leftarrow b-1$\;
      }
      \colorbox{yellow}{Attempt transmission of 1 packet;}\\
      \If{collision}{
        %\tcc{Random backoff.}
        $r \leftarrow r+1$\;
        $s \leftarrow \min (s+1,S)$\;
        $b \leftarrow \mathcal{U}[0, 2^s {\rm{CW}_{min}} -1]$\;
      }
    }
    $r \leftarrow 0$\;
    \colorbox{yellow}{$s \leftarrow 0$;}\\
	 \eIf{success}{
      %\tcc{Random backoff.}
      \colorbox{yellow}{$b \leftarrow \mathcal{U}[0,2^{s}{\rm{CW}_{min}}-1]$;}\\
    }
    {
      Discard packet\;
      $b \leftarrow \mathcal{U}[0,2^s {\rm{CW}_{min}}-1]$\;
    }
  }
  Wait until there is a packet to transmit\;
}
\caption{CSMA/CA}
\label{alg:CSMA_CA}
\end{algorithm}

\begin{algorithm}[ht!!!]
\While{the device is on}
{
  $r \leftarrow 0$ ; $s \leftarrow 0$\;
  $b \leftarrow \mathcal{U}[0,2^s{\rm{CW}_{min}}-1]$\;
  \While{there is a packet to transmit}{
    %\tcc{Initialize $a$.}
    \Repeat{($r = R$) or (success)}{
      %\tcc{First, backoff.}
      \While{$b>0$}{
        wait 1 slot\;
        $b \leftarrow b-1$\;
      }
      \colorbox{yellow}{Attempt transmission of 1 packet;}\\
      \If{collision}
      {
        %\tcc{Random backoff.}
        $r \leftarrow r+1$\;
        $s \leftarrow \min (s+1,S)$\;
        $b \leftarrow \mathcal{U}[0, 2^s {\rm{CW}_{min}} -1]$\;
      }
    }
    $r \leftarrow 0$\;
    \colorbox{yellow}{$s \leftarrow 0$;}\\
    \eIf{success}{
      %\tcc{Random backoff.}
      \colorbox{yellow}{$b \leftarrow (2^{s}{\rm{CW}_{min}})/2-1$;}\\
    }
    {
      Discard packet\;
      $b \leftarrow \mathcal{U}[0, 2^s {\rm{CW}_{min}}-1]$\;
    }
  }
  Wait until there is a packet to transmit\;
}
\caption{Basic CSMA/ECA}
\label{alg:CSMA_ECA}
\end{algorithm}

\begin{algorithm}[ht!!!]
\While{the device is on}
{
  $r \leftarrow 0$ ; $s \leftarrow 0$\;
  $b \leftarrow \mathcal{U}[0,2^s{\rm{CW}_{min}}-1]$\;
  \While{there is a packet to transmit}{
    %\tcc{Initialize $a$.}
    \Repeat{($r = R$) or (success)}{
      %\tcc{First, backoff.}
      \While{$b>0$}{
        wait 1 slot\;
        $b \leftarrow b-1$\;
      }
      \colorbox{yellow}{Attempt transmission of 1 packet;}\\
      \If{collision}{
        %\tcc{Random backoff.}
        $r \leftarrow r+1$\;
        $s \leftarrow \min (s+1,S)$\;
        $b \leftarrow \mathcal{U}[0, 2^s {\rm{CW}_{min}} -1]$\;
      }
    }
    $r \leftarrow 0$\;
    %$s \leftarrow 0$\;
    \eIf{success}{
      %\tcc{Random backoff.}
      \colorbox{yellow}{$b \leftarrow (2^{s}{\rm{CW}_{min}})/2-1$;}\\
    }
    {
      Discard packet\;
      $b \leftarrow \mathcal{U}[0, 2^s {\rm{CW}_{min}}-1]$\;
    }
  }
  Wait until there is a packet to transmit\;
}
\vspace{0.2cm}
\caption{CSMA/ECA with hysteresis}
\label{alg:CSMA_ECA_hist}
\end{algorithm}

\begin{algorithm}[ht!!!]
\While{the device is on}
{
  $r \leftarrow 0$ ; $s \leftarrow 0$\;
  $b \leftarrow \mathcal{U}[0,2^s\rm{CW}_{min}-1]$\;
  \While{there is a packet to transmit}{
    %\tcc{Initialize $a$.}
    \Repeat{($r = R$) or (success)}{
      %\tcc{First, backoff.}
      \While{$b>0$}{
        wait 1 slot\;
        $b \leftarrow b-1$\;
      }
      \colorbox{yellow}{Attempt transmission of $2^s$ packets;}\\
      \If{collision}{
        %\tcc{Random backoff.}
        $r \leftarrow r+1$\;
        $s \leftarrow \min (s+1,S)$\;
        $b \leftarrow \mathcal{U}[0, 2^s  \rm{CW}_{min} -1]$\;
      }
    }
    $r \leftarrow 0$\;
    %$s \leftarrow 0$\;
    \eIf{success}{
      %\tcc{Random backoff.}
      \colorbox{yellow}{$b \leftarrow (2^{s}\rm{CW}_{min})/2-1$;}\\
    }
    {
      Discard packet\;
      $b \leftarrow \mathcal{U}[0, 2^s \rm{CW}_{min}-1]$\;
    }
  }
  Wait until there is a packet to transmit\;
}
\vspace{0.2cm}
\caption{CSMA/ECA with hysteresis and fair-share}
\label{alg:CSMA_ECA_hist_fair}
\end{algorithm}

@@

An example of the operation of CSMA/ECA with hysteresis is presented in Fig. \ref{fig:csma_eca_different_backoff} for a network of $4$ contending stations.
For each station the backoff counter value at each station is indicated in every slot and the transmissions are represented by rounded rectancles. \Az{In the first slot shown, all stations are in the first backoff stage ($s = 0$, $\text{CW}_{\min} = 16$) and have chosen a random backoff value $b \sim \mathcal{U}[0,15]$. Stations $3$ and $4$ happen to have the same backoff counter value, and when their counter reaches zero, their transmissions collide. Therefore, they increase their backoff stage ($s = 1$) and select another random backoff value accordingly ($b \sim \mathcal{U}[0,31]$). On the other hand, stations $1$ and $2$ manage to successfully transmit their packets and move to the deterministic behavior for their next transmission, both selecting a deterministic value equal to $7 = 2^s \rm{CW}_{min}/2 - 1$. The first slot in which all stations have different backoff values, marked in the figure with a blue dashed line, can be viewed as the beginning of the collision-free operation. Since in this case no two stations will transmit simultaneously, there will be no collision and, by switching to the deterministic behavior, they will all periodically transmit in the same relative position. Note that stations $3$ and $4$ move to deterministic behavior when they are in backoff stage $s=1$. Therefore, they both use $15 = 2^s \rm{CW}_{min}/2 - 1$ as their deterministic backoff value and send two packets at each transmission opportunity (fair-share). The length of the deterministic cycle for the network in this example is $32$ slots, the least common multiple of all stations' cycles.}
% The stations initially pick a random backoff.}
% Those stations with the same backoff value collide and choose a new backoff value.
% When all the stations have a different backoff value, no more collisions will occur.
% At this point the behaviour of the system is cyclic and deterministic.
% Note that the state of the system at the beginning of the cycle is the backoff counter of the four stations: 4, 9, 14 and 2.
% It is exactly the same state that the first slot of the next cycle.

% The length of the cycle (32 slots) is indicated by the arrows at the bottom of the figure.
% The stations that collide double their contention window, the deterministic backoff after successes and the number of transmitted packet in each transmission attempt.

\begin{figure*}[!!!htb]
\centering
\includegraphics[width=7.0in]{figures/csma_eca_different_backoff}
\caption{An example of contention among CSMA/ECA stations with hysteresis and fair-share.}
\label{fig:csma_eca_different_backoff}
\end{figure*}

@@

%\AzDel{In the present workshop paper, we focus on a simple greenfield scenario with
%a single traffic class, ideal channel, ideal clocks, and saturation conditions.
%We are currently evaluating more complex scenarios that we outline in Section \ref{sec:future_work}.} \AzCom{Remove?! Can't really fit it in!}

In this section, the performance of the different CSMA protocols is evaluated through simulations. The presented results show the saturation throughput and fairness (Jain's Fairness Index (JFI) \cite{jain1984quantitative}) when the number of contenders increases. Additionally, the evolution of the probability that a slot contains a collision is plotted to provide further insights on the operation of each protocol, and specially whether the collision-free operation is reached.

@@

To evaluate the four protocols, a network of $N$ nodes is considered, where each node is within the coverage area of the others. The channel does not introduce errors and the nodes are set to be in saturation (always have packets to transmit). The number of contenders, $N$, ranges from $2$ to $50$, and one thousand instances of the simulation are performed for each $N$. All plotted results show the 95\% confidence intervals.

@@

In Figure \ref{Fig:throughput}, the throughput of all four schemes is plotted. As can be observed, the throughput for CSMA/CA decreases with the number of nodes, as the binary exponential backoff reduces the number of transmission attempts to keep the collisions low. The basic CSMA/ECA is able to achieve the collision-free operation if the number of nodes is lower than the deterministic cycle length, i.e., $N\leq \text{CW}_{\min}/2$ \AzCom{This time the $-1$ is not required, but it should be cycle length instead of backoff value, right?}. This is why in this figure, a phase transition is observed at $N=8$. When $N$ is larger than $\text{CW}_{\min}/2=8$, although the collision-free operation is not possible, the throughput obtained remains higher than in CSMA/CA. This is because, regardless of whether or not the collision-free operation is reached, in CSMA/ECA, a node always has a lower collision probability immediately after a successful transmission, without affecting the collision probability of other nodes.

@@

Clearly, reaching a collision-free operation is highly desirable, but not always possible in the basic CSMA/ECA. CSMA/ECA with Hysteresis exactly addresses this issue by simply not resetting the backoff stage after a successful transmission. In other words, the deterministic backoff value ($2^s \text{CW}_{\min}/2-1$) is chosen based on the backoff stage, $s$, in which the transmission was successful. This way, a larger number of contenders can reach a collision-free operation (Figure \ref{Fig:throughput}), but at the expense of a lower throughput for $N\leq 8$ compared to the basic CSMA/ECA. This is because if collision-free operation can be reached with a lower deterministic backoff value, there will be fewer empty slots, hence higher channel access efficiency. Furthermore, CSMA/ECA with hysteresis has a lower long-term fairness (Figure \ref{Fig:fairness}) compared to the basic CSMA/ECA or even the legacy CSMA/CA protocol, as once the network has reached the collision-free state, nodes that use a large backoff stage will have fewer transmission opportunities than others.

To address the aforementioned fairness issue when hysteresis is used, CSMA/ECA with hysteresis and fair-share was introduced, \Az{whereby stations that access the channel less frequently use packet aggregation to send more (precisely, $2^s$) packets per attempt.} As observed in Figure \ref{Fig:throughput}, using aggregation we are able not only to provide fair access, but also to significantly increase the throughput. {The throughput increases with the number of nodes because for larger $N$, the collision-free operation is reached at higher backoff stages, which means more packets will be aggregated and transmitted at every attempt, thus improving the efficiency of the network.}

@@

\section{Conclusions and Future Work}
\label{sec:future_work}

In this paper we have described a suitable replacement of the prevalent CSMA/CA protocol used in the IEEE 802.11 standard. Compared to the CSMA/CA protocol, the proposed protocol, called CSMA/ECA with Hysteresis and Fair-Share, offers performance boost in terms of throughput, while preserving long-term fairness. In fact, under certain conditions, collision-free operation can be reached in a distributed and adaptive manner, even as the number of nodes in the network increases. Furthermore, this protocol is designed with backward compatibility in mind, so that it can operate with legacy nodes without degrading their performance. All these properties make the proposed protocol a good candidate to replace CSMA/CA in the upcoming revisions of the standard.

\Az{We have evaluated the protocols under discussion in a very simple scenario using ideal channel, ideal clocks, and in saturation conditions. Furthermore, all the participating stations use the same protocol and have a single traffic  class. Our proposed protocol is based on CSMA/ECA, which is shown \cite{?} to work well in non-ideal conditions, and, although not presented in this manuscript, simulation results confirm that CSMA/ECA with hysteresis and fair-share inherits the flexibility and robustness of its predecessors.

The proposed protocol has yet to be implemented in real prototypes. However, current IEEE 802.11 commercial network interfaces do not allow for setting the backoff to a deterministic value after successful transmissions, despite recent efforts on creating firmware that enables higher degree of manipulation  \cite{tinnirello2012wmp}. An alternative is using radio-frequency identification (RFID) devices, where there is a programmable hardware that allows for implementing arbitrary protocols. RFID uses also a contention protocol which is prone to collisions, and we believe it can be a first arena in which we will be able to develop working prototypes to validate some of the ideas proposed in the present paper.}

%\section{Future Work}
